```{r setup, cache=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(fig.path="./figures/")
```

---
title: "Reproducible Research: Week 1 Project"
output: html_document
keep_md: true
---

## Loading Environment Packages/Etc.
In the code below, I am loading up environment variables and packages to ensure that the code can be run
irrespective of the particular modules and packages that are already loaded on my machine.
```{r environment and library loading, warning=FALSE, message=FALSE}
is.installed <- function(mypkg){
  is.element(mypkg, installed.packages()[,1])
}

if (!is.installed("data.table")){
  install.packages("data.table")
}

if (!is.installed("lubridate")){
  install.packages("lubridate")
}

if (!is.installed("ggplot2")){
  install.packages("ggplot2")
}

if (!is.installed("lattice")){
  install.packages("lattice")
}

if (!is.installed("knitr")){
  install.packages("knitr")
}

library(data.table)
library(lubridate)
library(ggplot2)
library(lattice)
library(knitr)
```

## Loading and preprocessing the data
In this step, I am loading up the data and assuming that it has already been unpacked via ZIP extraction or 
some other mechanism. The assignment did not require us to add code to specifically unzip the file.  
**Note** The code to deal with whether a given day of the week is a weekend or a weekday is used from the start.
```{r load and process data, warning=FALSE, message=FALSE}
activityData <- read.csv('activity.csv', stringsAsFactors = FALSE)
activityData$date <- as.POSIXct(as.Date(activityData$date), format = "%Y-%M-%D")
activityFrame <- data.frame(Date = activityData$date,
                              WeekDay = tolower(weekdays(activityData$date)),
                              Steps = activityData$steps,
                              interval = activityData$interval
                            )

# Code below determines weekday/weekend distinction 
activityFrame$DayType = ifelse(activityFrame$WeekDay == "saturday" | activityFrame$WeekDay == "sunday", "Weekend", "Weekday")
rm(activityData) # <-- Removing the data to save some space
```

# Calculate Total Number of Steps Per Day
In this block, I calculate the total number of steps per day and print the result to the resulting 
document. My original code generates PNG files; however, in this case, I chose to print the 
screen for the purpose of using RMarkdown and Knitr
```{r calculate steps per day, warning=FALSE, message=FALSE}
stepsByDay <- aggregate(activityFrame$Steps, by=list(activityFrame$Date), FUN = sum, na.rm = TRUE)
names(stepsByDay) <- c("Day", "TotalSteps")
p1 <- hist(stepsByDay$TotalSteps, 
     xlab = "Total Steps", 
     ylab = "Frequency of Steps", 
     main = "Histogram - Total Steps By Day",
     ylim = range(1:30),
     col = "Light Grey"
)
```

```{r workaround_plot1, warning=FALSE, echo=FALSE, message=FALSE}
p1
```

# Calculate Mean/Median Of Steps Per Day
Below we extract the mean and median steps per day as requested and print them to the screen
```{r print mean/median steps per day, warning=FALSE, message=FALSE}
stepsmean <- mean(stepsByDay$TotalSteps)
stepsmedian <- median(stepsByDay$TotalSteps)
```
**Mean:** `r round(stepsmean, digits=2)`  
**Median:** `r stepsmedian`  

# What is the average daily activity pattern?
The code below creates a plot describing the daily activity pattern
```{r daily activity pattern plot, warning=FALSE, message=FALSE}
intervalTimeSeries <- aggregate(activityFrame$Steps, by=list(activityFrame$interval), FUN = mean, na.rm = TRUE)
names(intervalTimeSeries) <- c("Interval","AverageSteps")
p2 <- plot(x = intervalTimeSeries$Interval,
     y = intervalTimeSeries$AverageSteps,
     lwd = 4,
     type = "l",
     xlab = "Interval", 
     ylab = "Average Steps", 
     main = "Time Series -- Average Steps By Interval",
     col = "Black"
)
```
`r p2`

## Find The Interval With Largest Max Value
Below I find the interval that has the largest max average steps value and print to the screen.
```{r calculate which interval has the largest max value, warning=FALSE, message=FALSE}
# Find The Interval Corresponding To Max Mean
maxPosition <- which(intervalTimeSeries$AverageSteps == max(intervalTimeSeries$AverageSteps))
intervalValue = intervalTimeSeries[maxPosition, 1]
```
**Interval With Largest Max Value:** `r intervalValue`

## Imputing missing values
The code below is used to identify missing values and imput them
```{r calculate and imput the missing values, warning=FALSE, message=FALSE}
# Find Number Of NA Values
naCount <- sum(is.na(activityFrame$Steps))

# Get Positions of NA Values
naPositions <- which(is.na(activityFrame$Steps))

# Imputed Values Creation
replacementVector <- rep(mean(activityFrame$Steps, na.rm=TRUE), times=length(naPositions))
activityFrame[naPositions, "Steps"] <- replacementVector
```

## Now plot everything again, this time with the values there.
The plot below looks at the total steps using the same basic histogram design, but
this time with the imputed data
```{r plot3, warning=FALSE, echo=FALSE}
# Now compute all the numbers again
revisedStepsByDay <- aggregate(activityFrame$Steps, by=list(activityFrame$Date), FUN = sum, na.rm = TRUE)
names(revisedStepsByDay) <- c("Day", "TotalSteps")
p3 <- hist(revisedStepsByDay$TotalSteps, 
     xlab = "Total Steps", 
     ylab = "Frequency of Steps", 
     main = "Histogram - Total Steps By Day, with Imputed Data",
     ylim = range(1:40),
     col = "Light Grey"
)
```

```{r workaround_plot3, warning=FALSE, echo=FALSE, message=FALSE}
p3
```

## What Are The Revised Mean And Median
After we have replaced the NA with our calculated value, what are the values? The code
below calculates those values
```{r revised_mean_median, echo=FALSE, warning=FALSE}
revised_stepsmean <- mean(revisedStepsByDay$TotalSteps)
revised_stepsmedian <- median(revisedStepsByDay$TotalSteps)
```
**Revised Mean:** `r format(revised_stepsmean, digits=2)`    
**Revised Median:** `r format(revised_stepsmedian, digit=2)`   

## Activity Pattern Differences Between Weekends and Weekdays
Below I do some processing to come up with the steps by weekend/weekday. Also, for some clarity and ease of plotting,
I created two sub-tables "weekend" and "weekday" that allow for very straightforward plotting using R's base
plotting system
```{r make panel plot for weekday/weekend, warning=FALSE, message=FALSE}
# Now plot the average across day type
daytypeTable <- aggregate(activityFrame$Steps, 
                          by=list(activityFrame$DayType, activityFrame$WeekDay, activityFrame$interval), 
                          FUN = mean, na.rm = TRUE)

names(daytypeTable) <- c("DayType", "DayOfWeek", "Interval", "AverageSteps")
weekend <- subset(daytypeTable, DayType == "Weekend")
weekday <- subset(daytypeTable, DayType == "Weekday")

par(mfrow=c(2,1))
p4 <- plot(weekend$Interval, weekend$AverageSteps, type="l", lwd=2, 
            col="Black", xlim=range(0:2500), ylim=range(0:300),
            xlab="Interval", ylab="Average Steps",
            main="Average Steps Over Intervals -- Weekend Days"
)

p4 <- p4 + plot(weekday$Interval, weekday$AverageSteps, type="l", lwd=2, 
                col="Black", xlim=range(0:2500), ylim=range(0:300),
                xlab="Interval", ylab="Average Steps",
                main="Average Steps Over Intervals -- Weekday Days"
)
```

`r p4`
